---
title: "Introduction to R: The Basics"
author: "Rony Rodriguez-Ramirez"
date: "2024-02-01"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Introduction

Welcome to this introductory tutorial on R! R is a powerful language and environment for statistical computing and graphics. It offers a wide variety of statistical and graphical techniques and is highly extensible. One of the key features of R is its package ecosystem, the most notable of which is the tidyverse, a collection of packages designed for data science.

This tutorial is designed for individuals with minimal statistical background. We will cover the basics of R, including how to install and load packages, import data, perform simple data manipulations, and create basic visualizations.

## Installing R and RStudio

Before we begin, you need to have R and RStudio installed on your computer. R is the underlying statistical computing environment, while RStudio provides a convenient and powerful user interface.

1. Download R from [The Comprehensive R Archive Network (CRAN)](https://cran.r-project.org).
2. Download RStudio from [RStudio's website](https://rstudio.com/products/rstudio/download/).

## Getting Started

### Setting Up Your Environment

Open RStudio and let's get started by installing and loading the tidyverse package:

```{r,eval=FALSE}
install.packages("tidyverse")
library(tidyverse)

``` 

The first line will install the package, and the second will load the package to your current R session. After you install your package, you do not need to use the first line again in your scripts.

## Assignment Operator (`<-`)

The assignment operator `<-` is used to assign values to variables in R. It is considered more readable and is preferred over the `=` operator in the R community. There are debates regarding when we should use `=` instead of `<-` but I am leaving that out of this tutorial for now.

### Example

```{r}
# Assigning value 10 to variable x
x <- 10
x
```

## Understanding Data Types in R

R supports various data types, each serving different purposes in data analysis. Familiarity with these data types is essential for effective data manipulation. Below are the primary data types you'll encounter in R:

### Numeric

Numeric data types are used to represent real numbers. They can be integers (without decimal points) or doubles (with decimal points), though R treats all numbers as floating-point values by default.

```{r}
my_integer <- 42  # An integer
my_double  <- 42.5 # A double
```

### Character

Character data types represent text. In R, text values are enclosed in quotes.

```{r}
my_character <- "Hello, world! I love econometrics!"
my_character
```

### Logical

Logical (or boolean) data types represent TRUE or FALSE values. They are the result of conditions or logical operations.

```{r}
my_logical <- TRUE
is_greater <- (5 > 2) # Evaluates to TRUE

my_logical
is_greater
```

### Factor

Factors are used to represent categorical data. They are especially useful for representing groups or levels (e.g., "Low", "Medium", "High").

```{r}
my_factor <- factor(c("Low", "Medium", "High", "Low", "Medium"))

my_factor
```

### Date and POSIXct

Date and POSIXct represent dates and date-times, respectively. They are crucial for time series analysis or any analysis that involves dates.

```{r}
my_date     <- as.Date("2024-01-01")
my_datetime <- as.POSIXct("2024-01-01 12:00:00")

my_date
my_datetime
```

### Vectors

Vectors are basic data structures in R that contain elements of the same type. Use the `c()` function to create a vector:

```{r}
my_vector <- c(1, 2, 3, 4, 5)
my_vector
```

### List

Lists in R can contain elements of different types, including numbers, strings, vectors, and even other lists.

```{r}
my_list <- list(
  name   = "Eric Taylor", 
  scores = c(90, 85, 88), 
  passed = TRUE
)

my_list
```

Each type has its specific purpose and a set of functions that operate on it, making it essential to choose the right type for your data analysis tasks.

### Data Frame and Tibble

A data frame is a table-like structure that can hold columns of different data types. A tibble is a modern take on the data frame, part of the `tidyverse`, designed to be more user-friendly. We will discuss more about this is the data cleaning section.

```{r}
my_data_frame <- data.frame(
  id = 1:3,
  name = c("Emma", "Rony", "Jostin"),
  score = c(90, 85, 88)
)

my_tibble <- tibble(
  id = 1:3,
  name = c("Emma", "Rony", "Jostin"),
  score = c(90, 85, 88)
)
```

### Basic R Operations

Now, let's look at the basic operations. R can be used as a simple calculator. Try typing `2 + 2` in the console. You can assign values to variables using `<-`.

```{r}
x <- 2 + 2
x
```

## Importing Data

R can read data from various sources. The `read_csv()` function from the readr package (part of the `tidyverse`) is commonly used for reading CSV files:

```{r,eval = FALSE}
my_data <- read_csv("path/to/your/file.csv")
```

Replace `"path/to/your/file.csv"` with the actual file path to a csv file.

## Pipe operator

The pipe operator `%>%`, heavily used in tidyverse packages, allows you to pass the result of one expression as the first argument to the next expression. It's a powerful tool for chaining together a sequence of operations in a clear and concise manner. Instead of nesting functions or using intermediate variables, you can use the pipe operator to form a pipeline of operations.

### How It Works

Consider a simple example without the pipe operator:

```{r}
result <- sum(c(1, 2, 3, 4, 5))
result
```

With the pipe operator, the same operation looks like this:

```{r}
c(1, 2, 3, 4, 5) %>%
  sum() 
```

Although this is a simplistic example, it illustrates how `%>%` passes the left-hand side of the operator as the first argument to the function on the right-hand side. This becomes particularly useful with functions from the `dplyr` package for data manipulation.


::: {.callout-tip title="Remember!"}
"Piping" in R can be seen as "chaining." This means that we are invoking multiple method calls.

```{r}
#| eval: false
# Pipe
rony %>% 
  wake_up(time = "5:30") %>% 
  get_out_of_bed() %>% 
  do_exercise() %>% 
  shower() %>% 
  get_dressed() %>% 
  eat(meal = "breakfast", coffee = TRUE) %>% 
  brush_teeth() %>%
  work(effort = "mininum")

# No pipe
  work(
    brush_teeth(
      eat(
        get_dressed(
          shower(
            do_exercise(
              get_out_of_bed(
                wake_up(rony, time = "5:30")
              ), 
            )
          )
        ), meal = "breakfast", coffee = TRUE
      )
    ), effort = "minimum"
  )
```
:::

## Data Manipulation with `dplyr`

The `dplyr` package offers a set of functions for manipulating data frames. You do not have to load this package since it is already loaded within the tidyverse package.

Let's look at the following functions:

- `filter()`: Extracts a subset of rows based on conditions.
- `select()`: Selects columns by name.
- `mutate()`: Creates new columns or modifies existing ones.
- `summarize()`: Summarizes multiple values into a single value.

Image that I would like to have a filtered dataset that contains observations with scores higher than 85. I can run the following:

```{r}
filtered_data <- my_data_frame %>%
  filter(score > 85) %>%
  select(name, score)
```

In this example:

1. `my_data_frame` is passed as the first argument to the `filter()` function, which filters rows where `score > 85`.
2. The result of `filter()` is then passed as the first argument to `select()`, which keeps only the columns name and score.
3. The final result is assigned to `filtered_data`.

Benefits of Using the Pipe Operator

- **Readability**: The flow of operations is from left to right, similar to how we read text, making the code easier to follow.
- **Maintainability**: It's easier to add or remove steps in the data processing pipeline without having to rewrite function calls or manage temporary variables.
- **Clarity**: Each step in the pipeline can be clearly seen, making it easier to understand what each part of the code is doing.

The pipe operator is a cornerstone of the tidyverse approach to data manipulation, enabling clear and expressive code that closely resembles natural language. As you become more familiar with R and the tidyverse, you'll find the pipe operator indispensable for efficient data analysis and transformation.

### Mutate, Summarize, and other Functions

The `mutate()` function allows you to create new columns in your data frame or change existing ones based on some operations or functions applied to the data. It's particularly useful for feature engineering, calculating new metrics, or preparing your data for analysis.

Let's say we have a data frame `my_data_frame` that contains student IDs, names, and scores. We want to add a new column that shows whether each student passed based on their score, with a score of 85 or above considered a pass.

```{r}
# Add a new column 'passed' with boolean values: TRUE if score >= 85, FALSE otherwise
my_data_frame <- my_data_frame %>%
  mutate(passed = score >= 85)

my_data_frame
```

This will add a new column passed to my_data_frame, where each row will have `TRUE` if the student's score is 85 or higher, and `FALSE` otherwise.

Suppose you want to standardize the scores (subtract the mean and divide by the standard deviation) for analysis.

```{r}
# Standardize the 'score' column
my_data_frame <- my_data_frame %>%
  mutate(
    score_standardized = (score - mean(score)) / sd(score)
  )

my_data_frame

```

This operation creates a new column `score_standardized` where each student's score is standardized.

### Using Multiple mutate() Operations

You can chain multiple operations within a single `mutate()` call. For example, if you wanted to add a column for the grade (A, B, C, etc.) based on the standardized score, you could do it within the same `mutate()`:

```{r}
# Add a grade column based on standardized scores
my_data_frame <- my_data_frame %>%
  mutate(
    score_standardized = (score - mean(score)) / sd(score),
    grade = case_when(
      score_standardized >= 1  ~ "A",
      score_standardized >= 0  ~ "B",
      score_standardized < 0   ~ "C"
    )
  )

my_data_frame
```

Here, `case_when()` is used to assign grades: "A" for standardized scores >= 1, "B" for scores >= 0 and < 1, and "C" for scores < 0. This illustrates how `mutate()` can be used to perform multiple transformations simultaneously, enhancing the readability and efficiency of your data manipulation code.

## Summarizing Data with `summarize()`

The `summarize()` function in `dplyr` is used to create summary statistics for a dataset or subsets of a dataset. This function can be particularly powerful when combined with `group_by()`, which groups the data frame by one or more variables.

### Basic Usage of `summarize()`

Here's how you can use `summarize()` to calculate the average (mean) score of all students in `my_data_frame`:

```{r}
average_score <- my_data_frame %>%
  summarize(
    mean_score = mean(score)
  )

average_score
```

This code calculates the mean of the `score` column and stores the result in a new data frame called `average_score` with a single column `mean_score`.

Now, image that you have some observations with missing values for score. In that case, you would like to ignore those missing values when estimating the mean score. Therefore, instead of using `mean(score)` you can use the argument `na.rm = TRUE`, as follows: `mean(score, na.rm = TRUE)`.

### Grouped Summaries

To calculate the average score by a categorical variable (e.g., by a hypothetical `class` column), you first group the data by that variable using `group_by()`, then apply `summarize()`:

First, let's add a hypothetical class variable to our data frame to categorize students into two classes, "A" and "B". We'll use the mutate() function to accomplish this:

```{r}
my_data_frame <- my_data_frame %>%
  mutate(
    class = if_else(id %% 2 == 0, "A", "B")
  )

my_data_frame
```

This code adds a new column `class` to `my_data_frame`, where students with an even id are assigned to class "A" and students with an odd id are assigned to class "B".

Now that we have a class variable, we can calculate the average score for each class:

```{r}
average_score_by_class <- my_data_frame %>%
  group_by(class) %>%
  summarize(
    mean_score = mean(score)
  )

average_score_by_class
```

This will calculate the mean score for each class separately.

### Multiple functions within a `summarize`.

You can also use `summarize()` to apply multiple summary functions at once. For example, to calculate the mean, median, and standard deviation of scores:

```{r}
summary_statistics <- my_data_frame %>%
  summarize(
    mean_score = mean(score),
    median_score = median(score),
    sd_score = sd(score)
  )

summary_statistics
```

### Using `summarize()` with `mutate()`

While `summarize()` reduces your data to a single summary row, `mutate()` can be used alongside to add summary columns to the original data frame. For example, adding a column with the mean score to each row:

```{r}
my_data_frame_with_mean <- my_data_frame %>%
  mutate(mean_score = mean(score))

my_data_frame_with_mean
```

## Subset Operators

R provides several operators for subsetting objects like vectors, lists, and data frames.

### The `[ ]` Operator

The `[ ]` operator is used for extracting elements from vectors, matrices, or data frames based on their indices or conditions.

#### Example with Vector

```{r}
# Creating a vector
vec <- c(1, 2, 3, 4, 5)

# Subsetting the second element
vec[2]
```

#### Example with Data Frame

```{r}
# Creating a data frame
df <- data.frame(Name = c("Alice", "Bob", "Charlie"), Age = c(25, 30, 35))

# Subsetting rows
df[1:2, ]
```

### The `[[ ]]` Operator

The `[[ ]]` operator is used to extract elements from a list or a single element from a data frame, allowing for nested subsetting in lists.

#### Example with List

```{r}
# Creating a list
lst <- list(name = "Alice", age = 25, scores = c(90, 80, 85))

# Extracting a single element
lst[[3]]
```

### The `$` Operator

The `$` operator is used for extracting elements by name from lists or data frames, making it particularly useful for accessing columns in a data frame.

#### Example with Data Frame

```{r}
# Accessing a column by name
df$Name
```

## Getting Help

If you don't understand a function, you can use the question mark and the function you would like to know more about, like this:

```{r, eval = FALSE}
?mutate()
```